---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# PooledCohort

<!-- badges: start -->
<!-- badges: end -->

The goal of `PooledCohort` is to give researchers who study risk prediction for cardiovascular disease (CVD) a unified interface to implement the Pooled Cohort Equations, the PREVENT equations, and other contemporary CVD risk calculators 

*Why use the Pooled Cohort Risk prediction equations?*

The 2017 American College of Cardiology and American Heart Association blood pressure guideline recommends using 10-year predicted atherosclerotic cardiovascular disease risk to guide the decision to initiate or intensify antihypertensive medication. The guideline recommends using the Pooled Cohort Risk prediction equations to predict 10-year atherosclerotic cardiovascular disease risk. Therefore, a new method for predicting atherosclerotic cardiovascular disease risk should be evaluated based on how much it can improve risk prediction relative to the Pooled Cohort Risk prediction equations.

## Installation

<!-- You can install the released version of PooledCohort from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("PooledCohort") -->
<!-- ``` -->

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("bcjaeger/PooledCohort")
```
## Example

A basic example below computes 10-year atherosclerotic cardiovascular risk using the original Pooled Cohort Risk equations for a person who is black/white and male/female with 

- 55 years of age
- 213 mg/dL total cholesterol
- 50 mg/dL high density lipoprotein (HDL) cholesterol
- 120 mm Hg systolic blood pressure
- no antihypertensive medication use
- no current smoking
- no prevalent diabetes

First we will use a dataset that requires no modification of any variable to be plugged into `predict_10yr_ascvd_risk()`

```{r example}

library(PooledCohort)
library(dplyr, warn.conflicts = FALSE)

example_data <- data.frame(
  sex = c('female', 'female', 'male', 'male'),
  race = c('black', 'white', 'black', 'white'),
  age_years = rep(55, times = 4),
  chol_total_mgdl = rep(213, times = 4),
  chol_hdl_mgdl = rep(50, times = 4),
  bp_sys_mmhg = rep(120, times = 4),
  bp_meds = rep('no', times = 4),
  smoke_current = rep('no', times = 4),
  diabetes = rep('no', times = 4),
  stringsAsFactors = FALSE
)

example_data

```

A convenient way to use `predict_10yr_ascvd_risk()` is within `dplyr::mutate()`:

```{r}

example_risk <- example_data %>% 
  mutate(
    risk = predict_10yr_ascvd_risk(
      sex = sex,
      race = race,
      age_years = age_years,
      chol_total_mgdl = chol_total_mgdl,
      chol_hdl_mgdl = chol_hdl_mgdl,
      bp_sys_mmhg = bp_sys_mmhg,
      bp_meds = bp_meds,
      smoke_current = smoke_current,
      diabetes = diabetes
    )
  ) %>% 
  select(sex, race, risk)

example_risk

```

A similar interface is available for the PREVENT equations. Note how we set `race` to `NULL` (PREVENT equations do not use race), and add values for statin use, estimated glomerular filtration rate, and body mass index. 

```{r}

example_data %>% 
  mutate(
    risk = predict_10yr_ascvd_risk(
      sex = sex,
      race = NULL,
      age_years = age_years,
      chol_total_mgdl = chol_total_mgdl,
      chol_hdl_mgdl = chol_hdl_mgdl,
      bp_sys_mmhg = bp_sys_mmhg,
      bp_meds = bp_meds,
      smoke_current = smoke_current,
      diabetes = diabetes, 
      statin_meds = "no",
      egfr_mlminm2 = 100,
      bmi = 28,
      equation_version = "Khan_2023"
    )
  ) %>% 
  select(sex, race, risk)

```


## Data formatting

Data usually need to be modified slightly before being plugged into a Risk calculator. For example, instead of a `race` variable with values of `black` and `white`, the data may have a `race` variable with values of `african_american`, `caucasian`, and `other`.

```{r}

example_data_granular <- data.frame(
  sex = c('female', 'female', 'male', 'male'),
  race = c('african_american', 'caucasian', 'african_american', 'other'),
  age_years = rep(55, times = 4),
  chol_total_mgdl = rep(213, times = 4),
  chol_hdl_mgdl = rep(50, times = 4),
  bp_sys_mmhg = rep(120, times = 4),
  bp_meds = rep('no', times = 4),
  smoke_current = rep('no', times = 4),
  diabetes = rep('no', times = 4),
  stringsAsFactors = FALSE
)

```


While you can always modify variables in your data so that they meet the same format as the variables in `example_data` above, you may prefer to let `predict_10yr_ascvd_risk()` modify those variables for you: 

```{r}

# a mapping from the current race categories to 
# the 'black' and 'white' categories used by the
# Pooled Cohort Risk equations.

race_levels <- list(
  black = 'african_american',
  white = c('caucasian', 'other')
)

example_risk_granular <- example_data_granular %>% 
  mutate(
    risk = predict_10yr_ascvd_risk(
      sex = sex,
      race = race,
      race_levels = race_levels,
      age_years = age_years,
      chol_total_mgdl = chol_total_mgdl,
      chol_hdl_mgdl = chol_hdl_mgdl,
      bp_sys_mmhg = bp_sys_mmhg,
      bp_meds = bp_meds,
      smoke_current = smoke_current,
      diabetes = diabetes
    )
  ) %>% 
  select(sex, race, risk)

example_risk_granular

```

## A picky estimator

`predict_10yr_ascvd_risk()` is a picky estimator that will throw hard stops at you if it doesn't like the data you give it. In particular, if the input data has continuous variables with values outside the range of recommended values for the Pooled Cohort equations, you will get an error message.

```{r, error = TRUE}

predict_10yr_ascvd_risk(
  sex = 'male',
  race = 'black',
  age_years = 35, # age recommendation: 40-79
  chol_total_mgdl = 213,
  chol_hdl_mgdl = 55,
  bp_sys_mmhg = 120,
  bp_meds = 'no',
  smoke_current = 'no',
  diabetes = 'no'
)

```

This is meant to help you avoid mis-use of the Pooled Cohort Risk equations. However, if you _must_ go outside the range of recommended values, you can set the argument `override_boundary_errors` to `TRUE`.  

```{r}

predict_10yr_ascvd_risk(
  sex = 'male',
  race = 'black',
  age_years = 35,
  chol_total_mgdl = 213,
  chol_hdl_mgdl = 55,
  bp_sys_mmhg = 120,
  bp_meds = 'no',
  smoke_current = 'no',
  diabetes = 'no',
  override_boundary_errors = TRUE
)

```

